### 2013-03-28 doc/sosp13/intro.tex
intro/overview reorg

We believe we can support irregular applications with good performance and cost-effectiveness with commodity hardware for two main reasons. First, commodity multicore processors have become extremely fast with high clock rates, large caches and robust DRAM bandwidth. So with careful software engineering we can multiplex hundred of thousands of workers into a single core. Second, commodity networks offer high bandwidth as long as messages are large enough. So with enough data requests in flight simultaneously, the system can aggregate small messages into large enough ones.

In this paper we introduce \Grappa, a software runtime system that allows a commodity cluster of x86-based nodes connected via an InfiniBand network to be programmed as if it were a single, large, shared-memory NUMA (non-uniform memory access) machine with scalable performance for irregular applications. \Grappa leverages as much freely available and commodity infrastructure as possible. We use unmodified Linux for the operating system and an off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is used for process setup and tear down. GASNet~\cite{gasnet} is used as the underlying mechanism for remote memory reads and writes using active message invocations. \Grappa adds three main software components: (1) a \emph{lightweight tasking\/} layer that supports a context switch in as little as 38ns and distributed global load balancing; (2) a \emph{distributed shared memory\/} layer that supports normal access operations such as \emph{read\/} and \emph{write\/} as well as synchronizing operations such as \emph{fetch-and-add\/}~\cite{fetchandadd}; and (3) a \emph{communication\/} layer that combines short messages to achieve peak bandwidth on commodity networks. As we will show later, \Grappa can tolerate latencies way beyond that of the network. Therefore, \Grappa can afford to \emph{trade latency for throughput}. By {\em increasing\/} latency in key components of the system we are able to: increase effective random access memory bandwidth by delaying and aggregating messages; increase synchronization rate by delegating atomic operations to gatekeeper cores, even when referencing node-local global data; and improve load balance by tolerating the delays incurred when work-stealing.

>>>>

We believe we can support irregular applications with good performance and
cost-effectiveness with commodity hardware for two main reasons. First,
commodity multicore processors have become extremely fast with high clock
rates, large caches and robust DRAM bandwidth. So with careful software
engineering we can multiplex hundred of thousands of workers into a single
core. Second, commodity networks offer high bandwidth as long as messages are
large enough. Therefore with enough data requests in flight simultaneously,
the system can aggregate small messages into large enough ones.

We build on these two observations and develop \Grappa, a software runtime
system that allows a commodity cluster of x86-based nodes connected via an
InfiniBand network to be programmed as if it were a single, large,
shared-memory NUMA (non-uniform memory access) machine with scalable
performance for irregular applications. \Grappa tolerates communication
latencies with lightweight user-level tasking layer that supports a context
switch in as little as 38ns and can sustain a large number of active workers.
It bridges the commodity network bandwidth gap with a communication layer that
combines short messages originating from a large number of active workers
into larger ones.

### 2013-03-28 doc/sosp13/overview.tex
intro/overview reorg

@@

\Grappa (Figure~\ref{fig:grappa}) has three main software components:
\begin{description}

\item [Tasking system.] The tasking system supports lightweight multithreading
to tolerate communication latency and global distributed work-stealing (i.e.,
tasks can be stolen from any node in the system), which provides automated
load balancing. The scheduler oversubscribes to have more worker threads than
required for latency tolerance. By having at least several threads ready to
run at all times, the scheduler can prefetch a context data into cache to
lower the likelihood of costly main memory accesses during a context switch.

\item[Distributed shared memory.] The DSM system provides
fine-grain access to data anywhere in the system. It supports synchronization
operations on global data, explicit local caching of any memory in the system,
and support for operations on remote data.
The DSM system design relies on the lightweight tasking system and
communication layer in order to offer high aggregate random
access bandwidth for accessing remote data.

\item[Communication layer.] Modern commodity networks
support high bandwidth only for large messages. Since irregular applications
tend to need frequent communication of small requests, the main goal of our
communication layer is to aggregate small messages into large ones to better
exploit what the network can offer. It is largely invisible to the application
programmer.

% software developer but helps to improve network performance when applications read and write only small pieces of data.

@@

% \paragraph{Exploiting Latency Tolerance} As we will show later, Grappa can tolerate latencies way beyond that of the network. Therefore, Grappa can afford to \emph{trade latency for throughput\/}: by {\em increasing\/} latency in key components of the system we are able to increase our aggregate random access bandwidth (by delaying and aggregating messages), our synchronization bandwidth (by delegating operations to remote nodes), and our ability to improve load imbalance (work stealing increases latency).

% \TODO{check to see we actually explain these three somewhere.  Ie, that aggregating increases throughput and latency, that delegating synchronization increases the rate and the latency at which we can eg atomically increment, and that stealing work increases the latency of some tasks (eg when their data is largely at the originating node) but provides greater throughput overall.}

In the next three sections we describe both \Grappa's main components and how they are implemented.

>>>>

\Grappa leverages as much freely available and commodity infrastructure as
possible. We use unmodified Linux for the operating system and an
off-the-shelf user-mode InfiniBand device driver stack~\cite{OFED}. MPI is
used for process setup and tear down. GASNet~\cite{gasnet} is used as the
underlying mechanism for remote memory reads and writes using active message
invocations. To this infrastructure, \Grappa adds  three main software components, shown in Figure~\ref{fig:grappa}:

@@

\begin{description}

\item [Tasking system.] The tasking system supports lightweight multithreading
to tolerate communication latency and global distributed work-stealing (i.e.,
tasks can be stolen from any node in the system), which provides automated
load balancing. The scheduler oversubscribes to have more worker threads than
required for latency tolerance. By having at least four workers per processor
core ready to run at all times, the scheduler can prefetch a worker's state
into cache to lower the likelihood of costly main memory accesses during a
context switch.

\item[Distributed shared memory.] The DSM system provides fine-grain access to
data anywhere in the system. It supports normal access operations such as
\emph{read\/} and \emph{write\/} as well as synchronizing operations such as
\emph{fetch-and-add\/}~\cite{fetchandadd}. It also offers explicit local
caching of any memory in the system, and delegates that directly operate on
remote data. The memory model offered is similar to what underpins
C/C++~\cite{N2480,N2800}, so it is familiar to programmers. The DSM system
design relies on the lightweight tasking system and communication layer in
order to offer high aggregate random access bandwidth for accessing remote
data.

\item[Communication layer.] The main goal of our communication layer is to
aggregate small messages into large ones, which invisible to the application
programmer. Its interface is based on active messages~\cite{vonEicken92}.
Since aggregation and deaggregation of messages needs to be very efficient, we
made it highly concurrent with careful use of lock-free synchronization
operations.

@@

As we will show throughout the paper, \Grappa can tolerate latencies way beyond
that of the network. Therefore, \Grappa can afford to \emph{trade latency for
throughput}. By {\em increasing\/} latency in key components of the system we
are able to: increase effective random access memory bandwidth by delaying and
aggregating messages; increase synchronization rate by delegating atomic
operations to gatekeeper cores, even when referencing node-local global data;
and improve load balance by tolerating the delays incurred when work-stealing.
